Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-03-02 application.tex
revising

In the inner few AU of the MMSN, however, $\beta$ depends on $z$. 
Given a fixed $\khat$, one can have $\beta > \beta_\mathrm{crit}$ at
the optically thick midplane, but with increasing height the disk
becomes optically thin and $\beta<\beta_\mathrm{crit}$. Our analytical
criteria is not valid for variable $\beta$, and we must turn to
numerical solutions of the linear to determine whether or not the
fundamental mode operates. We do this in the next
section.

@@

We remark that it is possible to satisfy $\beta < \beta_\mathrm{crit}$
by choosing a sufficiently large $\khat$. We
illustrate this by plotting at each radius the value of $\khat$ such
that $\beta = \beta_\mathrm{crit}$. We thus expect that the VSI,
should it occur in the inner disk, would only occur on very small
scales. It is not clear if such modes are robust (see below).

>>>>

In the inner few AU of the MMSN, however, $\beta$ depends on $z$. 
Given a fixed $\khat$, one can have $\beta > \beta_\mathrm{crit}$ at
the optically thick midplane, but with increasing height the disk
becomes optically thin and $\beta<\beta_\mathrm{crit}$. Our analytical
criteria is not valid for variable $\beta$, and we must turn to
numerical solutions of the linear problem to determine whether or not
the VSI operates. We do this in the next section.

@@

We remark that it is possible to satisfy $\beta < \beta_\mathrm{crit}$
by choosing a sufficiently large $\khat$. We
illustrate this by plotting at each radius the value of $\khat$ such
that $\beta = \beta_\mathrm{crit}$. We thus expect that the VSI,
should it develop in the inner disk, would only occur on very small
scales. It is not clear if such modes are robust (see below).

### 2015-03-02 paper.tex
updated all figs
revising

\begin{abstract}
  Baroclinic astrophysical disks with equilibrium rotation profiles
  $\Omega$ that depends on the distance away from the disk midplane are
  potentially unstable to the Vertical Shear Instability (VSI). It is
  is important to understand the linear instability before
  considering the VSI as a hydrodynamic route to turbulent angular
  momentum transport in accretion disks. We study the linear,
  axisymmetric VSI in radially-local but vertically-global models
  of accretion disks. We include an energy equation with thermal
  relaxation on a characteristic timescale $t_c=\beta\Omega_k^{-1}$,
  where $\beta$ is a dimensionless parameter and $\Omega_k$ is the
  Keplerian rotation frequency, and explore how the VSI depends on the
  thermodynamic response  of the disk by varying $\beta$. We find
  increasing $\beta$ away from zero dramatically reduces VSI growth
  rates because this generally introduces a strongly stabilizing
  vertical entropy gradient. Our linear calculations are consistent
  with previous hydrodynamic simulations, which indicates the VSI 
  requires a short thermal relaxation timescale. For vertically  
  isothermal disks, we quantify `fast'
  thermal relaxation as $\beta < \epsilon|q|/(\gamma-1)$, where
  $\epsilon$ is the disk aspect-ratio, $q$ is the radial power-law
  index for the equilibrium temperature profile (which gives rise to vertical 
  shear) and $\gamma$ is the adiabatic index. We apply our linear
  results to a disk model of the Minimum Mass Solar Nebulae (MMSN), and
  find the MMSN is unstable to the VSI at tens of AU with a
  characteristic growth times of tens of local orbital periods. This 
  suggests the VSI is  dynamically important in the outer parts of protoplanetary disks. 
\end{abstract}

@@

This paper is organized as follows. In \S\ref{setup} we list the
governing equations and global equilibrium disk models under
consideration. We formulate the linear stability problem in
\S\ref{linear} and simplify it by making the radially-local
approximation. We study the linear problem using two approaches. In
\S\ref{analytical} we present an analytical discussion of vertically
isothermal disks. Our treatment is grossly simplified owing to a
number of approximations made, but closely parallels previous work and 
allows us to highlight the essential effects of vertical shear. In
particular, we quantify a characteristic thermal timescale below which
the VSI is expected to operate. In \S\ref{numerical} we numerically
solve the linearized fluid equations in the radially-local
approximation. We find basic agreement with our analysis, including
the aforementioned characteristic thermal timescale. In
\S\ref{application} we apply our linear models to the Minimum Mass
Solar Nebulae (MMSN) and find that the VSI may operate in the outer
parts of the MMSN. We summarize and discuss our results in
\S\ref{summary}, including caveats and possible directions for future
work.

>>>>

\begin{abstract}
  A robust hydrodynamic mechanism for turbulent angular momentum
  transport is important for cold accretion disk such as 
  protoplanetary disks. The vertical shear instability (VSI) has 
  recently been suggested to provide such a pathway, since it operates in
  disks with rotation profiles that depends on the height from the
  disk  midplane, which is generally expected in 
  astrophysical disks. However, the VSI requires a short thermal
  timescale. A quantitative assessment of the thermodynamic
  requirement for the VSI is thus desirable for application to
  realistic disks. To this end,  
  we generalize the linear, axisymmetric VSI in radially local but
  vertically global models of accretion disks to include an energy
  equation with a thermal relaxation term that restores the disk
  temperature $T$ over a characteristic timescale
  $t_c=\beta\Omega_k^{-1}$, where $\beta$ is a dimensionless parameter
  and $\Omega_k$ is the Keplerian rotation frequency. We explore how
  the VSI depends on the thermodynamic response  of the disk by
  varying $\beta$. We find increasing $\beta$ away from zero
  dramatically reduces VSI growth rates because this generally
  introduces a strongly stabilizing vertical entropy gradient. Our
  linear calculations are consistent with previous hydrodynamic
  simulations. For vertically isothermal disks, we quantify `fast' 
  thermal relaxation as $\beta < \epsilon|q|/(\gamma-1)$, where
  $\epsilon$ is the disk aspect-ratio, $q = \p\ln T/\p\ln r$  ($r$
  being the cylindrical radius from the central object), and 
  $\gamma$ is the adiabatic index. We apply our linear
  results to a disk model of the Minimum Mass Solar Nebulae (MMSN), and
  find the MMSN is unstable to the VSI at tens of AU with a
  characteristic growth times of tens of local orbital periods. This 
  suggests the VSI is dynamically important in the outer parts of
  protoplanetary disks.  
\end{abstract}

@@

This paper is organized as follows. In \S\ref{setup} we list the
governing equations and global equilibrium disk models under
consideration. We formulate the linear stability problem in
\S\ref{linear} and simplify it by making the radially-local
approximation. We study the linear problem using two approaches. In
\S\ref{analytical} we present an analytical discussion of vertically
isothermal disks. Our treatment is grossly simplified owing to a
number of approximations made, but closely parallels previous work and 
allows us to highlight the essential effects of vertical shear. In
particular, we quantify a characteristic thermal timescale below which
the VSI is expected to operate. In \S\ref{numerical} we numerically
solve the linearized fluid equations in the radially-local
approximation, but relax all other assumptions made in our analytical approach.   
We find basic agreement with our analysis, including
the aforementioned characteristic thermal timescale. In
\S\ref{application} we apply our linear models to the Minimum Mass
Solar Nebulae (MMSN) and find that the VSI may operate in the outer
parts of the MMSN. We summarize and discuss our results in
\S\ref{summary}, including caveats and possible directions for future
work.
